### 2014-10-29 src/docs/talks/paris-dauphine-2014/stan-overview.tex
updated overview talk slides

\usepackage[
  scale=0.875,
  stdmathitalics=true,
  stdmathdigits=true]{lucimatx}
\linespread{1.02}
%\usepackage{helvet}

@@

\begin{document}
\sf%
\vspace*{6pt}
%
\noindent
\spc{\huge\bfseries \color{MidnightBlue}{Stan}}
\\[8pt]
\spc{\Large\bfseries \color{MidnightBlue}{Probabilistic Programming Language}}
\\
\vfill
\noindent
\spc\spc{\slshape\footnotesize Core Development Team:}
\\[2pt]
\spc\spc{\small Andrew Gelman, \  \myemph{Bob Carpenter}, \  Matt Hoffman}
\\
\spc\spc{\small Daniel Lee, \  Ben Goodrich, \  Michael Betancourt, \ }
\\
\spc\spc{\small Marcus Brubaker, \   Jiqiang Guo, \  Peter Li, \ }
\\
\spc\spc{\small Allen Riddell, \  Marco Inacio}
\vfill
\mbox{ } 
\hfill
\includegraphics[width=27.5pt]{img/stanlogo-main.pdf}\hspace*{9pt}
\\
\spc{\footnotesize Paris-Dauphine 2014}
\hfill 
{\footnotesize \tt mc-stan.org}

@@

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
\vspace*{-12pt}
\begin{itemize}\footnotesize
\item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
\end{itemize}
\item Stan \myemph{program}
\vspace*{-4pt}
\begin{itemize}\small
\item declares variables
\item codes log posterior (penalized likelihood)
\end{itemize}
\item Stan \myemph{inference}
\vspace*{-4pt}
\begin{itemize}\small
\item MCMC for full Bayesian inference
\item MLE for point estimation
\end{itemize}
\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
\\
{\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost C++ lib, googletest}
\end{itemize}

\sld{Who's Using Stan?}
\begin{itemize}
\item 750+ registrations for mailing list across the physical,
biomedical, and social sciences, plus engineering, finance and marketing

\item Subjects of Papers on Google Scholar in the sciences
\\
{\footnotesize clinical drug
trials, general computational statistics, entomology, opthalmology,
neurology, sociology and population dynamics, genomics, agriculture,
psycholinguistics, molecular biology, population dynamics, materials
engineering, botany, astrophysics, oceanography, election prediction,
fisheries, cancer biology, public health and epidemiology, population
ecology, collaborative filtering for recommender systems, climatology,
educational testing, natural language processing}
\end{itemize}

\sld{Books and Model Sets}
%
\begin{itemize}
\item 400+ page core language tutorial manual
\item separate installation, getting started, interface manuals 
\\ {\footnotesize  (RStan, PyStan, CmdStan)}
\item BUGS and JAGS examples (all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Types of scaling
\vspace*{-4pt}
\begin{itemize}\small
\item more data
\item more parameters
\item \myemph{more complex models}
\end{itemize}
\item for MCMC, measure (as function of scaling)
\vspace*{-4pt}
\begin{itemize}\small
\item time to convergence
\item time per effective sample at mixing
\item memory usage
\end{itemize}
\end{itemize}

\sld{Stan vs. BUGS / JAGS}
\begin{itemize}
\item JAGS is the former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
\\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
\item Stan (2.0) vs. BUGS on scaling IRT model
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item Stan about \myemph{10 times faster}
\item Stan memory and time scales linearly
\item linearly scales with number of question answers
\end{itemize}
\vspace*{-3pt}
\vfill
\hspace*{-12pt}
\includegraphics[height=0.6in]{img/irt_scale-eps-converted-to.pdf}
\includegraphics[height=0.45in]{img/irt-memory-2.png}
\item Stan can handle problems that choke BUGS and JAGS
\end{itemize}

@@

\begin{itemize}
\item \myemph{Platforms}
\\
Linux, Mac OS X, Windows
\item \myemph{C++ API}
\\
{\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
\item \myemph{Interfaces}
\vspace*{-4pt}
\begin{itemize}\footnotesize
\item \myemph{CmdStan}: Command-line or shell interface (direct executable)
\item \myemph{RStan}: R interface (Rcpp in memory)
\item \myemph{PyStan}: Python interface (Cython in memory)
\item \myemph{MStan$^*$}: MATLAB interface (lightweight external process)
\item \myemph{JuliaStan$^*$}: Julia interface (lightweight external process)
\end{itemize}
{\footnotesize ${}^*$ User contributed}
\end{itemize}
\vfill

@@

@@

\sld{Coding Probability Functions}
\begin{itemize}
\item \myemph{Vectorized} to allow scalar or container arguments
\\ {\footnotesize (containers all same shape; scalars broadcast as necessary)}
\item Avoid \myemph{repeated computations}, e.g. $\log \sigma$ in
\hspace*{-18pt}
{\small
\begin{eqnarray*}
\textstyle \log \, \mbox{\sf Normal}(y | \mu, \sigma)
& = & \textstyle \sum_{n=1}^N \log \, \mbox{\sf Normal}(y_n | \mu,\sigma)
\\[4pt]
& = & \textstyle \sum_{n=1}^N  - \log \sqrt{2\pi} \ - \log \sigma \ -
\frac{\textstyle y_n - \mu}{\textstyle 2\sigma^2}
\end{eqnarray*}
}
\item recursive \myemph{expression templates} to broadcast and cache scalars,
generalize containers (arrays, matrices, vectors)
\item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
  \sqrt{2 \pi}$) and calculate intermediate and return types
\end{itemize}

@@

\sld{L-BFGS Optimizer}
\begin{itemize}
\item Faster, lower memory, more stable than BFGS
\item Plug-and-play in interfaces
\vfill
\item Code and doc complete; under testing
\end{itemize}

@@

\sld{Thermodynamic Sampler}
\begin{itemize}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
\vfill
\item Prototype complete
\\ {\small (Betancourt paper on {\slshape arXiv}; for geometers)}
\end{itemize}

\sld{Marginal Maximum Likelihood}
\begin{itemize}
\item Marginalize out hierarchical parameters
\item Gradient-based nested optimization algorithm
\item Errors estimated via curvature as in MLE
\vfill
\item Design complete; awaiting parameter tagging
\end{itemize}

\sld{``Black Box'' VB and EP}
\begin{itemize}
\item \myemph{Black box} so can run any model
\\
{\footnotesize (Laplace or other approximations)}
\item Stochastic, data-streaming \myemph{variational Bayes} (VB)
\item Data-parallel \myemph{expectation propagation} (EP)
\\ {\footnotesize (cavity distributions provide general shard combination)}
\item Both use point estimates of parametric approximation to posterior
\item Optimize parameters to minimize KL divergence
\\
{\footnotesize (VB, EP measure divergence in opposite directions)}
\vfill
\item Prototype stage
\end{itemize}

@@

>>>>

\usepackage[
scale=0.875,
stdmathitalics=true,
stdmathdigits=true]{lucimatx}
\linespread{1.02}
% \usepackage{helvet}

@@

\begin{document}
\sf%
\vspace*{6pt}
% 
\noindent
\spc{\huge\bfseries \color{MidnightBlue}{Stan}}
\\[8pt]
\spc{\Large\bfseries \color{MidnightBlue}{Probabilistic Programming Language}}
\\
\vfill
\noindent
\spc\spc{\slshape\small Core Development Team:}
\\[2pt]
\spc\spc{\small Andrew Gelman, \  \myemph{Bob Carpenter}, \  Matt Hoffman}
\\
\spc\spc{\small Daniel Lee, \  Ben Goodrich, \  Michael Betancourt, \ }
\\
\spc\spc{\small Marcus Brubaker, \   Jiqiang Guo, \  Peter Li, \ }
\\
\spc\spc{\small Allen Riddell, \  Marco Inacio, \ Jeffrey Arnold, \ }
\\
\spc\spc{\small Mitzi Morris, \ Rob Trangucci}
\vfill
\mbox{ } 
\hfill
\includegraphics[width=27.5pt]{img/stanlogo-main.pdf}\hspace*{9pt}
\\
\spc{\small Stan 2.5.0 \ \footnotesize (October 2014)}
\hfill 
{\footnotesize \tt mc-stan.org}

@@

\begin{itemize}
\item Stan is an \myemph{imperative} probabilistic programming language
  \vspace*{-12pt}
  \begin{itemize}\footnotesize
  \item  cf., BUGS: declarative; \ Church: functional; \ Figaro: object-oriented
  \end{itemize}
\item Stan \myemph{program}
  \vspace*{-4pt}
  \begin{itemize}\small
  \item declares data and (constrained) parameter variables
  \item defines log posterior (or penalized likelihood)
  \end{itemize}
\item Stan \myemph{inference}
  \vspace*{-4pt}
  \begin{itemize}\small
  \item MCMC for full Bayesian inference
  \item MLE for point estimation
  \end{itemize}
\item Stan is \myemph{open source} \ {\footnotesize (BSD core C++, GPLv3 interfaces)}
  \\
  {\footnotesize hosted on GitHub; uses Eigen matrix lib, Boost C++ lib, googletest}
\end{itemize}

\sld{Who's Using Stan?}
\begin{itemize}
\item 850+ mailing list registrations;  5500 manual downloads (2.4.0)

\item Biological sciences: {\footnotesize
clinical drug trials, entomology, opthalmology,
neurology, genomics, agriculture, botany, fisheries,
cancer biology, epidemiology, population ecology, neurology
}
\item Physical sciences: {\footnotesize 
astrophysics, molecular biology, oceanography, climatology
}
\item Social sciences: {\footnotesize
 population dynamics, psycholinguistics, social networks, political science
}
\item Other: {\footnotesize materials engineering, finance, actuarial,
  sports, public health,
  collaborative filtering, educational testing}
\end{itemize}

\sld{Books and Model Sets}
% 
\begin{itemize}
\item 450+ page core language tutorial manual
\item Installation and getting started manuals by interface
  \\ {\footnotesize  (RStan, PyStan, CmdStan, MatlabStan, Stan.jl)}
\item BUGS and JAGS examples (most of all 3 volumes), 
\item Gelman and Hill, {\slshape Data Analysis Using Regression and
    Multilevel/Hierarchical Models}
\item Wagenmakers and Lee, {\slshape Bayesian Cognitive Modeling}
\item two other books in progress
\end{itemize}

\sld{Scaling and Evaluation}
\begin{itemize}
\item Types of scaling
  \vspace*{-4pt}
  \begin{itemize}\small
  \item more data
  \item more parameters
  \item \myemph{more complex models} \hfill (why we built Stan)
  \end{itemize}
\item for MCMC, measure \hfill (vs.\ BUGS / JAGS)
  \vspace*{-4pt}
  \begin{itemize}\small
  \item time to convergence \hfill (0.5--{\large$\infty$} faster)
  \item time per effective sample at mixing \hfill (ditto)
  \item memory usage \hfill (90\% less, linear scaling)
  \end{itemize}
\end{itemize}

\sld{Stan vs. JAGS (BUGS)}
\begin{itemize}
\item JAGS is the new BUGS, and former \myemph{state-of-the-art}, \myemph{black-box} MCMC method
  \\ {\footnotesize (Gibbs with slice sampling; chokes w.\ high posterior correlation)}
\item Stan (2.0) vs. BUGS on scaling multilevel 2PL-IRT model
  \vspace*{-4pt}
  \begin{itemize}\footnotesize
  \item Stan about \myemph{10 times faster}  ($n_{eff}/s$) and uses
    10\% memory
  \end{itemize}
  \vspace*{-3pt}
  \vfill
  \hspace*{-12pt}
  \includegraphics[height=0.6in]{img/irt_scale-eps-converted-to.pdf}
  \includegraphics[height=0.45in]{img/irt-memory-2.png}
\item Stan can handle problems that choke BUGS and JAGS
\end{itemize}

@@

\begin{itemize}
\item \myemph{Platforms}
  \\
  Linux, Mac OS X, Windows
\item \myemph{C++ API}
  \\
  {\footnotesize portable, standards compliant (C++03 now, moving to C++11)}
\item \myemph{Interfaces}
  \vspace*{-4pt}
  \begin{itemize}\footnotesize
  \item \myemph{CmdStan}: Command-line or shell interface (direct executable)
  \item \myemph{RStan}: R interface (Rcpp in memory)
  \item \myemph{PyStan}: Python interface (Cython in memory)
  \item \myemph{MatlabStan$^*$}: MATLAB interface (lightweight external process)
  \item \myemph{Stan.jl$^*$}: Julia interface (lightweight external process)
  \end{itemize}
  {\footnotesize ${}^*$ User contributed}
\end{itemize}
\vfill

@@

\sld{NUTS vs.\ Gibbs and Metropolis}

\includegraphics[width=0.9\textwidth]{img/nuts-vs.pdf}
\begin{itemize}
\item Two dimensions of highly correlated 250-dim distribution
\item 1M samples from Metropolis, 1M from Gibbs (thin to 1K)
\item 1K samples from NUTS, 1K independent draws
\end{itemize}

\sld{NUTS vs.\ Basic HMC}

\includegraphics[width=0.9\textwidth]{img/nuts-ess-1.pdf}
{\small
  \begin{itemize}
  \item 250-D normal and logistic regression models
  \item Vertical axis is effective sample size per sample (bigger better)
  \item Left) NUTS; \ \ Right) HMC with increasing $t = \epsilon L$
  \end{itemize}
}

\sld{NUTS vs.\ Basic HMC II}

\includegraphics[width=0.9\textwidth]{img/nuts-ess-2.pdf}

{\small
  \begin{itemize}
  \item Hierarchical logistic regression and stochastic volatility
  \item Simulation time $t$ is $\epsilon \ L$, step size ($\epsilon$)
    times number of steps ($L$)
  \item NUTS can beat optimally tuned HMC (latter very expensive)
  \end{itemize}
}

@@

\sld{Coding Probability Functions}
\begin{itemize}
\item \myemph{Vectorized} to allow scalar or container arguments
  \\ {\footnotesize (containers all same shape; scalars broadcast as necessary)}
\item Avoid \myemph{repeated computations}, e.g. $\log \sigma$ in
  \hspace*{-18pt}
  {\small
    \begin{eqnarray*}
      \textstyle \log \, \mbox{\sf Normal}(y | \mu, \sigma)
      & = & \textstyle \sum_{n=1}^N \log \, \mbox{\sf Normal}(y_n | \mu,\sigma)
      \\[4pt]
      & = & \textstyle \sum_{n=1}^N  - \log \sqrt{2\pi} \ - \log \sigma \ -
      \frac{\textstyle y_n - \mu}{\textstyle 2\sigma^2}
    \end{eqnarray*}
  }
\item recursive \myemph{expression templates} to broadcast and cache scalars,
  generalize containers (arrays, matrices, vectors)
\item \myemph{traits} metaprogram to \myemph{drop constants} (e.g., $-\log
  \sqrt{2 \pi}$ or $\log \sigma$ if constant) 
  and calculate intermediate and return types
\end{itemize}

@@

@@

\sld{Adiabatic Sampling}
\begin{itemize}
\item Physically motivated alternative to ``simulated''
  \myemph{annealing and tempering} (not really simulated!)
\item Supplies external \myemph{heat bath}
\item Operates through \myemph{contact manifold}
\item System relaxes more naturally between energy levels
  \vfill
\item Prototype complete
  \\ {\small (Betancourt paper on {\slshape arXiv}; for geometers)}
\end{itemize}

\sld{Maximum Marginal Likelihood}
\begin{itemize}
\item Fast, Approximate Inference
\item Marginalize out lower-level parameters
\item Optimize higher-level parameters and fix
\item Optimize lower-level parameters given higher-level
\item Errors estimated as in MLE
  \vfill
\item Design complete; awaiting parameter tagging
\end{itemize}

\sld{``Black Box'' VB}
\begin{itemize}
\item Fast, Approximate Inference
\item \myemph{Black box} so can run any model
  \\
  {\footnotesize (Laplace or other approximations)}
\item Stochastic, data-streaming \myemph{variational Bayes} (VB)
\item Optimize parameteric approximation to posterior to minimize KL
  divergence 
  \vfill
\item Prototype stage (collaborating with Alp Kucukelbir, Dave Blei,
  Rajesh Ranganath)
\end{itemize}

\sld{``Black Box'' EP}
\begin{itemize}
\item Fast, Approximate Inference
\item Data-parallel \myemph{expectation propagation} (EP)
  \\ {\footnotesize (cavity distributions provide general shard combination)}
\item Optimize parameteric approximation to posterior to minimize KL
  divergence 
  {\footnotesize (VB, EP measure divergence in opposite directions)}
  \vfill
  \vfill
\item Design stage (collaborating with Nicolas Chopin, Christian
  Robert, John Cunningham, Aki Vehtari, Pasi Jyl\"anki)
\end{itemize}

@@

\sld{``Naive Bayes'' Four Ways}
\begin{itemize}
%
\item Joint Distribution \ $p(\pi,\phi,z,w)$, defined by:
\vspace*{-3pt}
\begin{itemize}\small
\item $\pi \sim \mbox{\sf Dirichlet}(\alpha)$    \hfill (topic prevalence)
\item $\phi_k \sim \mbox{\sf Dirichlet}(\beta)$  \hfill (word
  prevalence in topic $k$)
\item $z_d \sim \mbox{\sf Categorical}(\pi)$     \hfill (topic for doc $d$)
\item $w_{d,n} \sim \mbox{\sf Categorical}(\phi_{z_d})$  \hfill (word
  $n$ in doc $d$)
\end{itemize}
%
\item Inference Problem \ $p(\tilde{z},\phi,\pi |  w, z, \tilde{w})$
\vspace*{-3pt}
\begin{itemize}\small
\item fully supervised learning: \ $p(\pi,\phi | w, z)$
\item semi-supervised learning: \ $p(\pi, \phi | w, z,
  \tilde{w})$
\item clustering: \ $p(\tilde{z} | \tilde{w})$
\item prediction: \ $p(\tilde{z} | w, z, \tilde{w})$
\end{itemize}
\end{itemize}
